---
title: 'Lab: visualizing CA COVID-19 data'
output: html_document
date: '2022-07-21'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.height = 5.5,
  fig.width = 8
)
```


```{r load libraries}
library(tidyverse)
```

```{r load data}
## you can downlaod the most up-to-date data
## and take a look at the data dictionary here: 
## https://data.ca.gov/dataset/covid-19-hospital-data1

ca_covid_hosp_data <- read_csv("https://raw.githubusercontent.com/isi-buds/program-2022/main/week2-material/day09/covid19hospitalbycounty.csv")
```

```{r visualize hospitalization trends of all counties}
ca_covid_hosp_data %>% 
  ggplot(aes(fill=county, x=todays_date, y=hospitalized_covid_confirmed_patients)) +
  geom_bar(position="stack", stat="identity")
```

The above plot tells us something about CA trends, but prevents us from comparing trends among counties, in addition to being absolutely hideous

```{r visualize hospitalization trends for 5 counties}
ca_five_county_covid_hosp_data <- ca_covid_hosp_data %>%
  filter(county %in% c("Orange", "Sacramento", "Santa Clara", "San Diego", "San Francisco"))

ca_five_county_covid_hosp_data %>% 
  ggplot(aes(fill=county, x=todays_date, y=hospitalized_covid_confirmed_patients)) +
  geom_bar(position="stack", stat="identity")
```

## Your task
The 5 county graph is more readable than the first graph, but still has tons of problems if one really wants to compare COVID-19 hospitalization trends across CA counties. Create your own visualization of the 5 county data, remembering best practices that we talked about in the lecture. There are more than one way of doing this, so don't be inhibited by trying to think of "the right" solution. 

## One possible solution

```{r define county color scheme}
county_colors <- c(Orange = "#E69F00", Sacramento = "#A857E6",`Santa Clara` = "#0072B2", `San Diego` = "#CC79A7", `San Francisco` = "#009E73")
```

```{r plot time series}
ca_five_county_covid_hosp_data %>% 
  ggplot(aes(x=todays_date, y=hospitalized_covid_confirmed_patients, group = county, color = county)) +
  geom_line(size = 1.3) + 
  scale_color_manual(name = "County", values = county_colors) + 
  theme_minimal(base_size = 15) + 
  labs(
    x = "Date",
    y = "Hospitalized COVID patients"
  ) + 
  ggtitle("CA County COVID-19 Hospitalization Trends")
  
```